---
title: "Colerectal Cancer"
author: "Yamuna Nunavath"
date: "2025-04-26"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.path = "figures/",    # all figures go into “figures/”
  dev = "png",              # or "pdf", "svg"…
  dpi = 300,                # good resolution
  fig.width = 6, fig.height = 4
)

options(repos = c(CRAN = "https://cran.rstudio.com"))

# Install and load necessary libraries
install.packages("ggplot2")
install.packages("dplyr")
install.packages("factoextra")
install.packages("cluster")
install.packages("caret")

library(caret)
library(ggplot2)
library(dplyr)
library(factoextra)
library(cluster)
library(reticulate)

# For biom, you might need to use reticulate to access Python's biom-format
biom <- import("biom")


```

```{r}
# Load the data (assuming data is already available in biom_data)
biom_data <- read.csv("/Users/yamunanunavath/Downloads/Data Harmonizing/kraken2_biom_data.csv")

# Rename columns for clarity
colnames(biom_data) <- c("OTU_ID", "Kraken2_Abundance")

# View the first few rows of the data
head(biom_data)
```

```{r}
# Log-transformation of Kraken2 abundance (log10)
biom_data$log_kraken2_abundance <- log10(biom_data$Kraken2_Abundance + 1)  # Adding 1 to avoid log(0)

# View first few rows of the log-transformed data
head(biom_data)

```

```{r}
# Plot distribution of Kraken2 abundance values
ggplot(biom_data, aes(x = Kraken2_Abundance)) +
  geom_histogram(binwidth = 50000, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Kraken2 Abundance Values", x = "Abundance", y = "Frequency") +
  theme_minimal()

# Plot distribution of log-transformed Kraken2 abundance values
ggplot(biom_data, aes(x = log_kraken2_abundance)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Log-Transformed Kraken2 Abundance Values", 
       x = "Log(Abundance)", y = "Frequency") +
  theme_minimal()


```




```{r}
# Calculate IQR to detect outliers
IQR_value <- IQR(biom_data$log_kraken2_abundance)
Q1 <- quantile(biom_data$log_kraken2_abundance, 0.25)
Q3 <- quantile(biom_data$log_kraken2_abundance, 0.75)

# Define the range for non-outlier values
lower_bound <- Q1 - 1.5 * IQR_value
upper_bound <- Q3 + 1.5 * IQR_value

# Identify outliers
outliers <- biom_data$log_kraken2_abundance[biom_data$log_kraken2_abundance < lower_bound | biom_data$log_kraken2_abundance > upper_bound]
outliers

```

6. Boxplot and Outlier Visualization:

```{r}
# Boxplot of log-transformed Kraken2 abundance
boxplot(biom_data$log_kraken2_abundance, main = "Log-Transformed Kraken2 Abundance", 
        ylab = "Log(Abundance)", col = "lightblue", border = "black", outline = TRUE)

# View outliers in terms of original Kraken2 abundance
outlier_OTUs <- biom_data[biom_data$log_kraken2_abundance %in% outliers, ]
head(outlier_OTUs)

# Boxplot for outlier OTUs
boxplot(outlier_OTUs$log_kraken2_abundance, main = "Outlier Kraken2 Abundance (Log-Transformed)",
        ylab = "Log(Abundance)", col = "lightgreen", border = "black", outline = TRUE)


```


7. Principal Component Analysis (PCA):

```{r}
# Perform PCA on multiple features (log-transformed abundance and other features if you have them)
pca_result <- prcomp(biom_data[, c("Kraken2_Abundance", "log_kraken2_abundance")], scale. = TRUE)

# PCA plot (PC1 vs PC2)
plot(pca_result$x[, 1], pca_result$x[, 2], 
     main = "PCA of Kraken2 Abundance", 
     xlab = "PC1", ylab = "PC2", col = "blue", pch = 19)



```




8. Clustering (K-means and Hierarchical Clustering):

```{r}

# Perform K-means clustering (3 clusters in this example)
set.seed(123)
kmeans_result <- kmeans(biom_data$log_kraken2_abundance, centers = 3)  # Adjust the number of clusters (k) if needed
biom_data$cluster <- as.factor(kmeans_result$cluster)

# Visualize K-means clustering result with PCA (PC1 vs PC2)
ggplot(biom_data, aes(x = pca_result$x[, 1], y = pca_result$x[, 2], color = cluster)) +
  geom_point() +
  labs(title = "PCA with K-means Clustering")

```



# Elbow Method for K-means Clustering

```{r}
# Calculate WSS for different values of k (Elbow Method)
wss <- numeric(10)
for (k in 1:10) {
  kmeans_result <- kmeans(biom_data$log_kraken2_abundance, centers = k)
  wss[k] <- kmeans_result$tot.withinss
}

# Plot the elbow method
plot(1:10, wss, type = "b", main = "Elbow Method for Optimal k", 
     xlab = "Number of Clusters", ylab = "Total Within-cluster Sum of Squares")

```


9. Statistical Tests (ANOVA, Kruskal-Wallis):

```{r}
# One-Way ANOVA to compare log-transformed Kraken2 abundance between clusters
anova_result <- aov(log_kraken2_abundance ~ cluster, data = biom_data)
summary(anova_result)

# Kruskal-Wallis test for comparing clusters
kruskal_test <- kruskal.test(log_kraken2_abundance ~ cluster, data = biom_data)
print(kruskal_test)


```

10. Post-hoc Analysis (Tukey HSD):

```{r}
# Tukey HSD post-hoc test
tukey_result <- TukeyHSD(anova_result)

# Display summary of Tukey HSD results
summary(tukey_result)

# Plot the Tukey HSD results
plot(tukey_result)



```



11. Clustering Visualization and Interpretation:

```{r}
# Visualize K-means clusters in 1D (density plot)
ggplot(biom_data, aes(x = log_kraken2_abundance, color = cluster)) +
  geom_density() +
  labs(title = "Density Plot of Log-Transformed Kraken2 Abundance by Cluster",
       x = "Log(Abundance)", y = "Density") +
  theme_minimal()

# Boxplot of `log_kraken2_abundance` for each cluster
ggplot(biom_data, aes(x = cluster, y = log_kraken2_abundance, fill = cluster)) +
  geom_boxplot() +
  labs(title = "Boxplot of Log-Transformed Kraken2 Abundance by Cluster", 
       x = "Cluster", y = "Log(Abundance)") +
  theme_minimal()
```


12. Cluster Profiling and Summary Statistics:

```{r}
cluster_summary <- biom_data %>%
  group_by(cluster) %>%
  summarise(
    mean_abundance = mean(log_kraken2_abundance, na.rm = TRUE),
    median_abundance = median(log_kraken2_abundance, na.rm = TRUE),
    sd_abundance = sd(log_kraken2_abundance, na.rm = TRUE),
    min_abundance = min(log_kraken2_abundance, na.rm = TRUE),
    max_abundance = max(log_kraken2_abundance, na.rm = TRUE)
  )

print(cluster_summary)


```


13. Visualizing Differences Between Clusters:

```{r}
# Violin plot to visualize the distribution of log_kraken2_abundance per cluster
ggplot(biom_data, aes(x = cluster, y = log_kraken2_abundance, fill = cluster)) +
  geom_violin() +
  labs(title = "Violin Plot of Log-Transformed Kraken2 Abundance by Cluster", 
       x = "Cluster", y = "Log(Abundance)") +
  theme_minimal()



```


```{r}
ggplot(biom_data, aes(x = log_kraken2_abundance, y = Kraken2_Abundance, color = cluster)) +
  geom_point() +
  labs(title = "Scatter Plot of Log-Transformed Kraken2 Abundance vs Original Kraken2 Abundance by Cluster")

```


14. Cluster Validation:

```{r}
# Perform PCA on the features
pca_res <- prcomp(biom_data[, c("Kraken2_Abundance", "log_kraken2_abundance")], scale. = TRUE)

# Silhouette score for clustering
library(cluster)

# Calculate silhouette score
silhouette_score <- silhouette(kmeans_result$cluster, dist(pca_res$x[, 1:2]))

# Plot silhouette score
plot(silhouette_score, main = "Silhouette Plot for K-means Clustering")

```


15. Random Forest:

```{r}
install.packages("randomForest")
library(randomForest)

# Ensure that the target variable 'cluster' is a factor
biom_data$cluster <- as.factor(biom_data$cluster)

rf_model <- randomForest(cluster ~ log_kraken2_abundance + Kraken2_Abundance, data = biom_data)

# Print the model summary
print(rf_model)

# Check the importance of the features
importance(rf_model)

# Plot the feature importance
varImpPlot(rf_model)
```


```{r}
library(caret)

# Define the training control
train_control <- trainControl(method = "cv", number = 10)

# Train the random forest model
rf_tuned <- train(cluster ~ log_kraken2_abundance + Kraken2_Abundance, 
                  data = biom_data, 
                  method = "rf", 
                  trControl = train_control, 
                  tuneGrid = expand.grid(mtry = c(1, 2)))  # Valid mtry values
print(rf_tuned)


```
```{r}
varImpPlot(rf_tuned$finalModel)

```

```{r}
svm_tuned <- train(cluster ~ log_kraken2_abundance + Kraken2_Abundance, 
                   data = biom_data, 
                   method = "svmRadial", 
                   trControl = train_control, 
                   tuneGrid = expand.grid(sigma = c(0.1, 0.5), C = c(1, 10)))
print(svm_tuned)

```


16. Cross_validation:

```{r}

# Define the cross-validation method
train_control <- trainControl(method = "cv", number = 10)  # 10-fold CV

# Fit the Random Forest model with cross-validation
rf_cv_model <- train(cluster ~ log_kraken2_abundance + Kraken2_Abundance, 
                     data = biom_data, 
                     method = "rf", 
                     trControl = train_control)

# Print the cross-validation results
print(rf_cv_model)

```

```{r}


```



